---
title: "Analysing CAGE data using cageopener"
author: 
- name: "Vivek Bhardwaj"
  affiliation: 
  - "Max Planck Institute of Immunobiology and Epigenetics, Stübeweg 51, 79108, Freiburg, Germany"
  - "Faculty of Biology, University of Freiburg, Schänzlestraße 1, 79104, Freiburg, Germany"
- name: "Thomas Manke"
  affiliation: 
  - "Max Planck Institute of Immunobiology and Epigenetics, Stübeweg 51, 79108, Freiburg, Germany"
  email: "manke@ie-freiburg.mpg.de"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Abstract
**cageopener** can be used to analyse sequencing data generated from conventional as well as modern CAGE protocols.
Here we show the usability of cageopener in analysis of MAPCap data.

**MAPCap** (Multiplexed Affinity Purification of Capped RNA) is a newly developed CAGE protocol that allows fast and 
accutate detection of transcription start sites and gene expression analysis of a set of multiplexed samples.

## Quick Start

Here we show a set of minimal steps for the processing of MAPCap data for the detection of TSS. Starting from the 
raw fastq files, we can either perform quality trimming using a standard program (eg. cutadapt/trimgalore) or simply 
begin the analysis by creating a `CapSet` object. The `CapSet` object takes raw/quality trimmed fastq file paths as input,
along with a simple dataframe (called `sampleInfo`) which contains demultiplexing barcodes as rownames and another column 
called `samples` which contain the corresponding sample names. The following steps creates the object, demultiplexes the fastq, maps them, filters them and detects the TSS.

```{r eval=FALSE}
# create directories
dir.create("01_fastq")
dir.create("01_fastq/raw")
dir.create("01_fastq/trimmed")
dir.create("02_splitting")
dir.create("03_mapping")
dir.create("04_removedup")
dir.create("05_tssCalling")

# load the package 
library(mapcapR)

# provide demultiplexing barcodes and sample names
idxlist <- c("TGGAAC", "GGTTAC")
fnames <- c("wt_rep1", "wt_rep2") 

# make the CapSet object


cs <- newCapSet("MAPCap", "paired", 
	    "01_fastq/raw/MAPCapH_R1.fastq.gz", 
	    "01_fastq/raw/MAPCapH_R2.fastq.gz",
	    sampleInfo = data.frame(row.names = idxlist, samples = fnames))

# trim sample barcodes from sequence
cs <- trimFastqIndex(cs, outdir = "01_fastq/trimmed")

# demultiplex fastq
cs <- demultiplex_fastq(cs, max_mismatch = 1, outdir = "02_splitting", nthreads = 10)

# map fastq
cs <- mapCaps(cs, subread_idx, outdir = "03_mapping", nthreads = 20, logfile = "03_mapping/subread_mapping.log")

# filter PCR duplicates
cs <- filterDuplicates(cs, outdir = "04_removedup")

# detect TSS
cs <- detect_TSS(cs, groups = c("wt", "wt"), outfile_prefix = "05_tssCalling")
```



## Help and citations

### How to get help

cageopener questions could be posted to the [Bioconductor support site](https://support.bioconductor.org), which serves as a searchable knowledge base of questions and answers.

Posting a question and tagging with “cageopener” will automatically send an alert to the package authors to respond on the support site. See the first question in the list of Frequently Asked Questions (FAQ) for information about how to construct an informative post.


### How to cite cageopener

Please cite cageopener as follows:



## Detailed description of the above workflow

### Creating a CapSet object

The workflow begins by creating an object of class `CapSet` using the function `newCapSet`. This contains information about the experiment method (CAGE, RAMPAGE or MAPCap), along with fastq type (`single` or `paired` end) and the path of the fastq files. It also contains a sampleInfo dataframe.

The sampleInfo is a dataframe containing the demultiplexing barcodes as rownames and the corresponding samplenames within a column 
called `samples`.

With this information, create the CapSet object as follows.

```{r eval=FALSE}
# provide demultiplexing barcodes and sample names
idxlist <- c("TGGAAC", "GGTTAC")
fnames <- c("wt_rep1", "wt_rep2") 

# make the CapSet object
cs <- newCapSet("MAPCap", "paired", 
	    "01_fastq/raw/MAPCapH_R1.fastq.gz", 
	    "01_fastq/raw/MAPCapH_R2.fastq.gz",
	    sampleInfo = data.frame(row.names = idxlist, samples = fnames))

```


### Trimming the barcodes

Experiments like MAPCap and RAMPAGE produce multiplexed fastq files with sample indicies and PCR barcodes attached to
the fastq sequence. The tool `trimFastqIndex` trims off these barcodes and attaches them in the header of the fastq files 
for further processing (i.e. sample de-multiplexing and PCR duplicate removal). The output is a modified CapSet object 
that stors informatin about the fastq file location.

```{r eval=FALSE}
cs <- trimFastqIndex(cs, outdir = "01_fastq/trimmed")
```

### demultiplexing the fastq

Trimmed fastq files from previous steps can now be demultiplxed to produce fastq files corresponding to each sample,
using the tool `demultiplex_fastq`. The result is a modified CapSet object that contains location of demultiplexed files
along with processing statistics.

```{r eval=FALSE}
# demultiplex fastq
cs <- demultiplex_fastq(cs, max_mismatch = 1, outdir = "02_splitting", nthreads = 10)
```

Optionally, we can skip the above process and perform post-mapping de-multiplexing on the BAM files (see below).

### Mapping the fastqs

The demultiplexed fastqs can now be mapped using the `mapCaps` function. This function is a wrapper over the `subjunc`
function from `Rsubread` package. It additionally performes sorting and collects mapping statistics of the mapped files, 
stored in the modified `CapSet` object.

In order to run the function we first create a subread index of our genome.

```{r eval=FALSE}
dir.create("genome_index")
Rsubread::buildindex(basename = "genome_index/dm6", reference = "/path/to/dm6/genome.fa")
```

We can now perform the mapping.

```{r eval=FALSE}
# provide location of a subread index file
subread_idx <- "/data/akhtar/bhardwaj/my_annotations/drosophila_dm6/subread_index/dm6"
# map fastq
cs <- mapCaps(cs, subread_idx, outdir = "03_mapping", nthreads = 20, logfile = "03_mapping/subread_mapping.log")
```

**Note:** Since fastq de-multiplexing is an optinal step, we can perform mapping directly on the multiplexed files
and perform the de-multiplexing post-mapping using the `splitBAM_byIndex` and `splitBAM_byRepindex` functions.

### (optional) Post mapping de-multiplexing

```{r eval=FALSE}
splitBAM_byIndex(bamFile, index_list, outfile_list, max_mismatch = 0, nthreads = 10)
```

### Filtering PCR duplicates

Experiments like MAPCap and RAMPAGE provide us a way to remove sequencing reads which are PCR duplicates from the 
mapped data. random UMIs are added to the read sequence for this purpose. In MAPCap, pre-designed random barcodes 
present in the oligos serve as the UMIs, while in RAMPAGE, the sequences used as RT-PCR primers are treated as 
pseudo-random barcodes. PCR duplicates are recognized by identifying reads that map to the same start and end site, 
and contain the same random UMI.

The function `filterDuplicates` removes these PCR duplicate sequences (keeping only one copy in these cases), and 
creates de-duplicated BAM files. It returns the modified `CapSet` object with de-duplication statistics.

```{r eval=FALSE}
# filter PCR duplicates
cs <- filterDuplicates(cs, outdir = "04_removedup")
```

### Detection of TSS

cageopener implements a new method of detection of transcription start sites, which is a adopted from recently described methods for peak calling and differential binding analysis methods. Genome is divided into 10 base-pair windows and the TSS are detected as the 
windows that show an X-fold enrichment over a local background of 2 kb. Multiple consicutively enriched windows are then merged to detect broad TSSs. The method works well with replicates.

This method is implemented in the function `detect_TSS`, which returns a modified `CapSet` object with TSS detection statistics.

```{r eval=FALSE}
# detect TSS
cs <- detect_TSS(cs, groups = c("wt", "wt"), outfile_prefix = "05_tssCalling")
```


## Plotting and QC

The `sampleInfo` field of the CapSet object stores information about the read numbers kept at each step of processing, this 
information can be easily plotted using the function ``

```{r eval=FALSE}

```

Similar to the FRIP (Fraction of Reads In Peaks) score, the fraction of total filtered reads that fall within the 
detected TSSs (FRIT) can be plotted using the function `plotFRIT`.

```{r eval=FALSE}

```

In case of well annotated genomes, one way to check the quality of TSS detection is to look at the fraction of detected
TSSs that fall close to an annotated TSS in the genome. The cumulative fraction can be plotted for each sample, which can 
be used to compare samples. This can be done using the function `plotPrecision`, which takes the *known* TSS annotations as 
a TxDB object.

```{r eval=FALSE}

```



## Differential TSS expression analysis

For the experiments with two or more groups, cageopener can also be used to perform differential TSS expression analysis for a group of samples. The requirements for differential TSS expression analysis is the same as that for differential expression analysis of RNA-Seq data. At least two or more biological replicates per group is required.

The functions `fit_diffTSS` and `detect_diffTSS` utilize edgeR to perform differential expression analysis.

```{r eval=FALSE}

```


### Using spike-In controls

For the differential TSS expression analysis, the function `fit_diffTSS` shown above utilize the [TMM method]()
for normalization of TSS counts. In some cases however, spike-in normalizations are preferred. cageopener provides
a way to perform spike-in normalization during differential expression analysis, through the functions ``.

```{r eval=FALSE}

```


## Additional useful functions

### sample information


### getting the gene counts








